---
title: "dataExploration_Jacob"
output: html_document
author: "Jacob Johnson"
date: "2024-12-11"
---


**- Questions to be answered**
5. Define which regions are most affected by certain types of weather? (Which most vulnerable to heat waves and extreme cold)
6. Do each of the above trends hold when pnumonia deaths (flu seasons) are removed?



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#getting libraries
library(readr)
library(tidyverse)
library(ggplot2)
library(dplyr)

#reading in data
data_main <- read.csv("Data Files/combined_TAVGxMortality_Data.csv")
```




### Analysis of which regions/cities are most affected by certain types of weather? (Heat Waves & Extreme cold)

**Specific question #5:** Define which regions are most affected by certain types of weather? (Which most vulnerable to heat waves and extreme cold)

```{r}
# Define thresholds for heat waves and extreme cold
heat_wave_threshold <- 85  # Threshold for heat wave (in degrees Fahrenheit)
extreme_cold_threshold <- 20  # Threshold for extreme cold (in degrees Fahrenheit)

# Filter data for heat waves, extreme cold, normal conditions(temperatures between the thresholds)
heat_wave_data <- data_main %>%
  filter(TMAX_Max >= heat_wave_threshold)

extreme_cold_data <- data_main %>%
  filter(TMIN_Min <= extreme_cold_threshold)

normal_data <- data_main %>%
  filter(TMAX_Max < heat_wave_threshold & TMIN_Min > extreme_cold_threshold)


# Calculate average and total deaths during heat waves by week
heat_wave_deaths <- heat_wave_data %>%
  group_by(region) %>%
  summarise(
    avg_deaths = mean(All.Deaths, na.rm = TRUE),
    total_deaths = sum(All.Deaths, na.rm = TRUE)) %>% 
  arrange(desc(total_deaths))

# Calculate average and total deaths during extreme cold by week
extreme_cold_deaths <- extreme_cold_data %>%
  group_by(region) %>%
  summarise(
    avg_deaths = mean(All.Deaths, na.rm = TRUE),
    total_deaths = sum(All.Deaths, na.rm = TRUE)) %>% 
  arrange(desc(total_deaths))
```


**Deaths during Heat Wave/Extreme Cold/General Conditions**
```{r}
# custom colors for each region
custom_colors <- c(
  "1" = "#1f77b4",
  "2" = "#ff7f0e",
  "3" = "#2ca02c",
  "4" = "#d62728",
  "5" = "#9467bd",  
  "6" = "#8c564b"   
)


# Summarize data by week and region (general)
data_summary <- data_main %>%
  group_by(region, Week) %>%
  summarise(All.Deaths = sum(All.Deaths, na.rm = TRUE))

# Plot multiple lines based on region with custom colors
ggplot(data_summary, aes(x = Week, y = All.Deaths, color = as.factor(region), group = region)) +
  geom_line(size = 1) +
  scale_color_manual(values = custom_colors) +
  theme_minimal() +
  labs(title = "All Deaths by Week and Region (General Conditions)",
       x = "Week",
       y = "All.Deaths",
       color = "Region")

#--------------------------------------------------
  
# Summarize data by week and region (heat wave)
data_summary_heat_wave <- heat_wave_data %>%
  group_by(region, Week) %>%
  summarise(All.Deaths = sum(All.Deaths, na.rm = TRUE))

# Plot multiple lines based on region with custom colors
ggplot(data_summary_heat_wave, aes(x = Week, y = All.Deaths, color = as.factor(region), group = region)) +
  geom_line(size = 1) +
  scale_color_manual(values = custom_colors) +
  theme_minimal() +
  labs(title = "All Deaths by Week and Region (During Heat Waves)",
       x = "Week",
       y = "All.Deaths",
       color = "Region")

#--------------------------------------------------

# Summarize data by week and region (extreme cold)
data_summary_extreme_cold <- extreme_cold_data %>%
  group_by(region, Week) %>%
  summarise(All.Deaths = sum(All.Deaths, na.rm = TRUE))

# Plot multiple lines based on region with custom colors
ggplot(data_summary_extreme_cold, aes(x = Week, y = All.Deaths, color = as.factor(region), group = region)) +
  geom_line(size = 1) +
  scale_color_manual(values = custom_colors) +
  theme_minimal() +
  labs(title = "All Deaths by Week and Region (During extreme cold)",
       x = "Week",
       y = "All.Deaths",
       color = "Region")
```


**Difference between mortality during heat waves/freezes compared to normal**
```{r}
# Calculate average mortality during normal conditions
normal_mortality <- normal_data %>%
  summarise(avg_deaths = mean(All.Deaths, na.rm = TRUE))

# Calculate average mortality during heat waves
heat_wave_mortality <- heat_wave_data %>%
  summarise(avg_deaths = mean(All.Deaths, na.rm = TRUE))

# Calculate average mortality during extreme cold
extreme_cold_mortality <- extreme_cold_data %>%
  summarise(avg_deaths = mean(All.Deaths, na.rm = TRUE))

# Calculate differences in mortality rates
mortality_diff <- data.frame(
  Condition = c("Normal", "Heat Wave", "Extreme Cold"),
  Avg_Deaths = c(normal_mortality$avg_deaths, heat_wave_mortality$avg_deaths, extreme_cold_mortality$avg_deaths)
)

mortality_diff <- mortality_diff %>%
  mutate(Difference = Avg_Deaths / normal_mortality$avg_deaths)

mortality_diff

# Visualizing the differences in mortality rates
ggplot(mortality_diff, aes(x = Condition, y = Difference, fill = Condition)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Difference in Mortality Rates During Extreme Weather Conditions",
       x = "Condition",
       y = "Difference in Average Deaths",
       fill = "Condition")
```


```{r}
# Calculate average mortality during normal conditions, grouped by city
normal_mortality <- normal_data %>% filter((!City == "Utica")) %>%
  group_by(City) %>%
  summarise(avg_deaths_normal = mean(All.Deaths, na.rm = TRUE))

# Calculate average mortality during heat waves, grouped by city
heat_wave_mortality <- heat_wave_data %>%
  group_by(City) %>%
  summarise(avg_deaths_heat_wave = mean(All.Deaths, na.rm = TRUE))

# Calculate average mortality during extreme cold, grouped by city
extreme_cold_mortality <- extreme_cold_data %>%
  group_by(City) %>%
  summarise(avg_deaths_extreme_cold = mean(All.Deaths, na.rm = TRUE))

# Combining the Mortality Rates data into one data frame
mortality_diff <- normal_mortality %>%
  left_join(heat_wave_mortality, by = "City") %>%
  left_join(extreme_cold_mortality, by = "City") %>%
  mutate(
    heat_wave_diff = ((avg_deaths_heat_wave - avg_deaths_normal) / avg_deaths_normal),
    extreme_cold_diff = ((avg_deaths_extreme_cold - avg_deaths_normal) / avg_deaths_normal))

mortality_diff

# Visualizing the Results + Reshape the data for plotting
mortality_diff_long <- mortality_diff %>%
  pivot_longer(cols = c(heat_wave_diff, extreme_cold_diff), names_to = "Condition", values_to = "Difference")

# Plot the differences in mortality rates by city
ggplot(mortality_diff_long, aes(x = City, y = Difference, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Percentage difference in Mortality Rates During Extreme Weather 
       Conditions by City",
       x = "City",
       y = "Percentage difference in Average Deaths",
       fill = "Condition") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

- **Key Findings**:
  - In total **Normal** and **Heat wave** average deaths are very similar **0.0017%** difference
  - In total **Extreme cold** average deaths is lower by around **22%** difference
  - The cities **Fresno, CA** & **Miami, FL** have no extreme cold deaths
  - **Utica** experiences a slight increase in average deaths during heat waves compared to normal conditions.
  - **Houston** sees a significant increase in average deaths during extreme cold conditions compared to normal conditions.

- **Conclusion**:
  - Region 2 (Cities - Houston, Miami, and Phoenix) has an average difference 
    factor of **129%** increase in average deaths during extreme cold conditions 
    compared to normal conditions.
  - Region 5 (Cities - Indianapolis, Colorado Springs, and Boston) has an average
    difference factor of **97%** decrease in average deaths during heat waves 
    compared to normal conditions. 


### Analysis of which trends still hold when pneumonia deaths are removed

**Specific question #6:** Do each of the above trends hold when pneumonia deaths (flu seasons) are removed?


```{r}
# Group by month and calculate the total pneumonia deaths
data_main %>%
  group_by(Month) %>%
  summarise(Total_Pneumonia_Deaths = sum(Pneumonia.and.Influenza.Deaths, na.rm = TRUE)) %>% 
  ggplot(aes(x = factor(Month), y = Total_Pneumonia_Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Total Pneumonia Deaths by Month",
       x = "Month",
       y = "Total Pneumonia Deaths") +
  scale_x_discrete(labels = month.abb)  # Use abbreviated month names```

# Group by month and calculate the total deaths
data_main %>%
  group_by(Month) %>%
  summarise(Total_Deaths = sum(All.Deaths, na.rm = TRUE)) %>% 
  ggplot(aes(x = factor(Month), y = Total_Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Total Deaths by Month",
       x = "Month",
       y = "Total Deaths") +
  scale_x_discrete(labels = month.abb)  # Use abbreviated month names```
```


```{r}
# Summarize total deaths by age group
age_groups <- c("X.1.year..all.cause.deaths.", "X1.24.years..all.cause.deaths.",
                "X25.44.years", "X45.64.years..all.cause.deaths.", "X65..years..all.cause.deaths.")

data_main %>% filter(!is.na(`All.Deaths`)) %>% group_by(City, Month) %>% pivot_longer(
  cols = age_groups,
  names_to = 'Age_Group',
  values_to = 'deaths',
) %>% filter(!is.na(deaths) ) %>% group_by(City, Age_Group) %>% mutate(
   deaths_sd = sd(deaths),
   deaths_mean = mean(deaths),
   deathsZ = (deaths - deaths_mean) / deaths_mean,
) %>% filter(!is.na(deaths) ) %>% group_by(Month, Age_Group) %>%
  mutate(
    deathsZ = mean(deathsZ)
  ) %>% filter(!is.na(deathsZ) ) %>%  
  ggplot(aes(x = factor(Month), y = deathsZ, fill = Age_Group )) + geom_bar(stat = 'identity', position = 'dodge') + labs(title = "Seasonal Mortality Anomaly by Age Group", y = "Deaths as % of mean", x = 'Month', fill = "Age Group") + scale_fill_discrete(labels = c("<1 year (all cause deaths)", "1-24 years (all cause deaths)",
                "25-44 years", "45-64 years (all cause deaths)", "65+ years (all cause deaths)"))


# Subtract Pneumonia.and.Influenza.Deaths from All.Deaths
data_main_excluding_pneumonia <- data_main %>%
  mutate(All.Deaths = All.Deaths - Pneumonia.and.Influenza.Deaths)

data_main_excluding_pneumonia %>% filter(!is.na(`All.Deaths`)) %>% group_by(City, Month) %>% pivot_longer(
  cols = age_groups,
  names_to = 'Age_Group',
  values_to = 'deaths',
) %>% filter(!is.na(deaths) ) %>% group_by(City, Age_Group) %>% mutate(
   deaths_sd = sd(deaths),
   deaths_mean = mean(deaths),
   deathsZ = (deaths - deaths_mean) / deaths_mean,
) %>% filter(!is.na(deaths) ) %>% group_by(Month, Age_Group) %>%
  mutate(
    deathsZ = mean(deathsZ)
  ) %>% filter(!is.na(deathsZ) ) %>%  
  ggplot(aes(x = factor(Month), y = deathsZ, fill = Age_Group )) + 
  geom_bar(stat = 'identity', position = 'dodge') + 
  labs(title = "Seasonal Mortality Anomaly by Age Group (Excluding Pneumonia and Influenza Deaths)",
       y = "Deaths as % of mean", x = 'Month',
       fill = "Age Group") + 
  scale_fill_discrete(labels = c("<1 year (all cause deaths)",
                                 "1-24 years (all cause deaths)",
                                 "25-44 years", "45-64 years (all cause deaths)",
                                 "65+ years (all cause deaths)"))
```

- **Key Findings**:
  - The total number of deaths ranges from **400,000-350,000** deaths a month
  - The total number of pneumonia & influenza deaths range from **30,000-16,000**
  - Using the seasonal mortality by age group chart compared to the same chart 
    excluding pneumonia & influenza deaths is little to no difference.

- **Conclusion**:
  - The amount of pneumonia & influenza deaths varies between age groups but the
    overall variation between the two charts is minimal and still shows that same
    trends in the data as before.
