---
output:
  html_document: default
  pdf_document: default
---

```{r}
library(tidyverse)
library(lubridate)
```


This is Nick trying to put together an example of how to arrange the weather data to compare to the deaths data.

```{r}
mortality <- read.csv("Deaths_in_122_U.S._cities_-_1962-2016._122_Cities_Mortality_Reporting_System_20241019.csv")
bostonWeather <- read_csv("Boston_USW00014739_1962-2016.csv")
```
## Aggregate Daily Weather Data To Weekly Averages

```{r}
# week <- as.Date(cut(bostonWeather$DATE, "week", start.on.monday = FALSE))
# weeklyBostonWeather <- aggregate(TAVG ~ week, bostonWeather, mean)
# 
# head(weeklyBostonWeather)

bostonWeather$DATE <- ymd(bostonWeather$DATE)

preparedBostonWeather <- bostonWeather %>% mutate(
  WEEK = week(DATE),
  Year = year(DATE),
  TAVG = (TMAX+TMIN)/2
) %>% 
  group_by(Year, WEEK)

weeklyBostonWeather <- aggregate(TAVG ~ (Year+WEEK), preparedBostonWeather, mean)
```

So we can easily cut the weather data down to weekly aggregates, some will be means while others sums or rates.

Below are box plots for yearly and weekly average temperature variance.

```{r}
preparedBostonWeather  %>%
  ggplot(aes(x = factor(Year), y = TAVG,)) + geom_boxplot()
```

```{r}

preparedBostonWeather %>% 
    ggplot(aes(x = factor(WEEK), y = TAVG)) + geom_boxplot()
```
## Joining the Two Datasets

Below I demonstrate a process for joining a city's mortality to its weather.

```{r}
bostonMortality <- mortality[mortality$City == 'Boston',]

```
```{r}
merge_list <- list(preparedBostonWeather, mortality)

by = join_by(Year, WEEK == WEEK)

# merge_list %>% reduce(full_join, by = c('Year', 'WEEK'))

joined <- left_join(bostonMortality, weeklyBostonWeather, by)

```

## Some Explorations

First, a basic scatterplot

```{r}

joined %>% ggplot(aes(x = TAVG, y = All.Deaths, color = month(mdy(Week.Ending.Date)))) + geom_point()


```

So, that isn't overly useful, we can see that there seems to be a negative correlation between Deaths and Temperature. Below I attempt to constrain it using a rolling average to establish a normal rate of mortality I can compare the weekly mortality to.

The method reveals seasonal effects on mortality.

```{r}
library(zoo)
```
```{r}
joined %>% mutate(
  normal_death_rate = rollmean(All.Deaths, 52*3, na.pad = TRUE)
) %>%
  ggplot(aes(x = TAVG, y = ((All.Deaths-normal_death_rate)/normal_death_rate), color = month(mdy(Week.Ending.Date)))) + geom_point()
```

```{r}

joined %>% mutate(
  normal_death_rate = rollmean(All.Deaths, 52*3, na.pad = TRUE)
) %>%
  ggplot(aes(x = TAVG, y = ((All.Deaths-normal_death_rate)/normal_death_rate))) + geom_point() + facet_wrap(~WEEK)


```

Below, I remove Pnumonia and Flu deaths from the sample to try and remove Flu season as a confounding factor.

```{r}
joined %>% mutate(
  sans_flu = (All.Deaths - Pneumonia.and.Influenza.Deaths),
  normal_death_rate = rollmean(sans_flu, 52*3, na.pad = TRUE)
) %>%
  ggplot(aes(x = TAVG, y = ((sans_flu-normal_death_rate)/normal_death_rate), color = month(mdy(Week.Ending.Date)))) + geom_point()
```

```{r}
joined %>% mutate(
  sans_flu = (All.Deaths - Pneumonia.and.Influenza.Deaths),
  normal_death_rate = rollmean(sans_flu, 52*3, na.pad = TRUE)
) %>%
  ggplot(aes(x = TAVG, y = ((sans_flu-normal_death_rate)/normal_death_rate))) + geom_point() + facet_wrap(~WEEK)

```

